\thetitle

### 2021-03-03 writing/rational-effort.tex
rational-*: rewording

In addition to success and failure information for the various trails, the
experiment software records the number of components a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as a metric.

A comparison of the effort of different modes of the rational programmer can
illuminate the comparison of the effectiveness of the three gradual typing
systems. For example, consider a debugging scenario.  If, for instance, both
Natural blame and Transient first blame trails are successful, the two modes of
the rational programmer can compete to see which one debugs the interesting
scenario with less effort. In general, if the effort distribution for a mode of
the rational programmer has a shorter tail and more volume around smaller values
compared to the effort distribution of another mode, then the first mode is
probably the more effective of the two.

Measuring effort can also tell whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.  Doing so means
comparing the effort distribution for a mode of the rational programmer with
another mode that ignores error information entirely and instead selects which
component to type next randomly.  Since each mutant has a finite number of
components, such random mode trails are naturally always successful. At the same
time though, the Natural, Transient, and Erasure random modes exhibit a normal
effort distribution across scenarios that should be distinct from the effort
distribution for all other modes of the rational programmer if the effectiveness
of the modes are not coincidental.

>>>>

In addition to success and failure information for the various trails, the
experiment records the number of components a rational programmer has
to equip with types along each trail ($\lvert \conf_n \setminus \conf_0
\rvert$). This number can serve as an additional metric to compare modes.

Comparing the effort of different modes of the rational programmer can
illuminate the comparitive effectiveness of the three gradual typing
systems. If, for instance, both
Natural blame and Transient first blame trails are successful for the same scenario, the two modes of
the rational programmer can compete to see which one debugs the
scenario with less effort. In general, if the effort distribution for a mode of
the rational programmer has a shorter tail and more volume around smaller values
compared to the effort distribution of another mode, then the first mode is
probably the more effective of the two.

Measuring effort can also reveal whether the observed effectiveness of the
rational programmer is an artifact of pure chance or not.
In particular, the effort distribution for a mode of the rational programmer can be compared with that of
the random mode that ignores error information entirely and instead selects which
component to type randomly.  Since each mutant has a finite number of
components, random mode trails are always successful. However, the random mode's effort
distribution will be thinly spread out across the range of trail lengths possible
in the set of debugging scenarios.
In contrast, the effort distribution of other modes should be quite different if their effectiveness is not coincidental.

### 2021-03-03 writing/rational-experiment.tex
restructuring benchmarks and moving earlier the experiment process figure
rational-*: rewording

Trails and their properties provide the tools for a rigorous examination of
blame for Natural, Transient and Erasure. In line with the discussion so far,
the experiment software collects data to answer three initial questions for 
interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of Natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and the experiment
uses the Natural blame and exception trails to answer it.

@@

In detail the answer to $Q_1$ demands a comparison of the success of the Natural
blame and Natural exception trails for all interesting debugging scenarios.  The
first step is to construct each mutant's scenario lattice and identify their
interesting debugging scenarios.  The experiment software extends the trails that start
from such roots according to the Natural-blame programmer.  If no scenarios can
be added to the trail, the experiment software checks whether the last scenario
of the trail type-checks or not. If it does, the software records that the
Natural-blame trail is successful; otherwise it is recorded as failing. The
process is repeated for the same roots again but for the other mode, Natural
exceptions.  Figure~\ref{fig:process} summarizes this experimental process for
one mode of the rational programmer and connects it with the mutations from
section~\ref{sec:mutate}.  After completing the experiment, the experiment
software reports the success/failure results of the trails to determine for each
root whether Natural blame is more useful than Natural exceptions.  Question
$Q_1$ has a positive answer if a root exists where the above is true because it
is evidence that there is at least one interesting scenario that the rational
programmer manages to debug with blame information.  The process is analogous
for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.

For $Q_*$, the process is a bit more involved. Answering this question calls for
a comparison of the percentage of scenarios where one mode is more useful than
the other and the inverse.  For instance, to decide whether blame for Natural is
more useful than first blame for Transient, the software compares the percentage
of interesting scenarios where Natural blame is more successful than Transient
first blame with the percentage of interesting scenarios where Transient first
blame is more successful than Natural blame.  Finally these steps are repeated
one more time to compare Natural blame and Transient-last blame and get a
complete picture of the comparative usefulness of blame in the two semantics.

\begin{figure}
  \centering
  \includegraphics[scale=0.36]{./Images/process}
  \caption{The experimental process for one mode of the rational
  programmer}
  \label{fig:process}
\end{figure}

>>>>

Trails and their properties provide the tools for a rigorous examination of
blame for Natural, Transient and Erasure. In line with the discussion so far,
the experiment collects data to answer three initial questions for 
interesting debugging scenarios:
\begin{itemize}
\item[$Q_1$] Is blame useful in the context of Natural?

@@

Table~\ref{fig:experiment-outline} summarizes how each question relates to
different kinds of trails/modes of the rational programmer. For example, experimental
question $Q_1$ asks whether blame is valuable for Natural and the experiment
uses the Natural blame and exception trails to answer it, so $Q_1$ shows up in the cells for Natural blame and Natural exceptions.

@@

In detail, the answer to $Q_1$ demands a comparison of the success of the Natural
blame and Natural exception trails for all debugging scenarios.  The
first step is to construct each mutant's scenario lattice and identify their
debugging scenarios.  The experiment then extends the trails that start
from those roots according to the Natural-blame programmer.  If no scenarios can
be added to the trail, the experiment checks whether the last scenario
of the trail type-checks or not. If it does, the
Natural-blame trail is successful; otherwise it is failing. 
Figure~\ref{fig:process} summarizes this experimental process for
one mode of the rational programmer and connects it with the mutations from
section~\ref{sec:mutate}.
The experiment repeats the process for the same roots with the other mode, Natural
exceptions.
After completing, the experiment
reports the success/failure results of the trails to determine the proportion of
scenarios where Natural blame is more useful than Natural exceptions.  Question
$Q_1$ has a positive answer if a root exists where the above is true because it
is evidence that there is at least one interesting scenario that the rational
programmer manages to debug with blame information.  The process is analogous
for $Q_2$ and $Q_3$, using the respective modes of the rational programmer.

\begin{figure}[bh]
  \centering
  \includegraphics[scale=0.36]{./Images/process}
  \caption{The experimental process for one mode of the rational
  programmer}
  \label{fig:process}
\end{figure}

For $Q_*$, the process is a bit more involved. Answering this question calls for
a comparison of the percentage of scenarios where one mode is more useful than
the other and the inverse.  For instance, deciding whether blame for Natural is
more useful than first blame for Transient requires comparing the percentage
of scenarios where Natural blame is more successful than Transient
first blame with the percentage where Transient first
blame is more successful than Natural blame. Repeating the whole process
for every pair of modes produces a
complete picture of the comparative usefulness of blame between all of the modes.

### 2021-03-03 writing/rational-transient.tex
challenges, start of rational: rewordings
rational-*: rewording

The transient semantics assigns blame to a set of components. In this setting, a
blame assignment has the following meaning: the witness value to the impedance
mismatch may have crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

@@

These two modes of rationalizing give rise to two different notions of trail.
\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component Transient adds to the blame set for $\conf$.
\end{quote}

and

@@

The definition of \emph{Transient exception trails} 
is analogous to that for Natural and, same as for Natural, we use it to  
specify when  
Transient-first and Transient-last blame are more useful than 
Transient exceptions for debugging a scenario.

>>>>

The Transient semantics assigns blame to a set of components. The blame
assignment says that the value witnessing the impedance mismatch may have
crossed the boundaries between elements in the set, and each crossing
checked the value's type in a shallow manner.

@@

These two modes of rationalizing give rise to two different notions of trail.

\begin{quote}
\it A \emph{Transient-first blame trail} is a sequence of scenarios
$\conf_0,...\conf_n$ of $\system$ where for all $0 \leq i \leq n - 1$,
$\conf_i \subset \conf_{i+1}$ and $\conf_{i+1} \setminus \conf_i =
\first{\mblame{\system, \conf_i}}$ where $\first{\mblame{\system, \conf}}$ is the
first component Transient adds to the blame set for $\conf$.
\end{quote}

@@

The definition of \emph{Transient exception trails} is analogous to that for
Natural. It is used to specify when Transient-first and Transient-last blame are
more useful than Transient exceptions for debugging a scenario.
